---
title: "Capstone_Main"
author: "Ty Hammond"
date: "2025-04-08"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load Libraries

```{r}
library(dplyr)
library(ggplot2)
library(readr)
library(tidyr)
library(caret)
library(tidyverse)
library(xgboost)
library(randomForest)
```

Load Dataset

```{r}
match_data <- read_csv("C:/Users/tyham/OneDrive - Colostate/Capstone_Project/match_data_eda.csv", show_col_types = FALSE)
```

*OVERALL DATA PREPROCESSING*

Ensure Date Column is Formatted Correct

```{r}
match_data$Date <- as.Date(match_data$Date, format = "%Y-%m-%d")
```

Handle Missing Data

```{r}
# Already Know whats missing from EDA
# Only drops 2 rows
match_data <- match_data %>%
  drop_na(`B365>2.5`, `B365<2.5`)
```

Convert Categorical Data

```{r}
match_data <- match_data %>%
  mutate(
    FTR = as.factor(FTR),
    OU_2.5 = as.factor(OU_2.5),
    HomeTeam = as.factor(HomeTeam),
    AwayTeam = as.factor(AwayTeam),
    Div = as.factor(Div)
  )
```


Data Preprocessing & Standardization

```{r}
# num cols
numeric_features <- c("B365H", "B365D", "B365A",
                      "imp_prob_H", "imp_prob_D", "imp_prob_A",
                      "norm_prob_H", "norm_prob_D", "norm_prob_A")

# standardize
preproc <- preProcess(match_data[, numeric_features], method = c("center", "scale"))

match_data_scaled <- match_data
match_data_scaled[, numeric_features] <- predict(preproc, match_data[, numeric_features])
```

*Rolling Team Features*
To get a measure of a teams history to get their form to predict matches

Create a Home and Away Dataset, also renaming for clarity
Doing this because teams play significantly different home vs away so makes modeling better

```{r}
# home
home_results <- match_data_scaled %>%
  select(Date, HomeTeam, FTHG, FTAG, HS, AS, HST, AST, HC, AC, HY, AY, HR, AR, FTR) %>%
  rename(
    Team = HomeTeam,
    Goals_For = FTHG,
    Goals_Against = FTAG,
    Shots_For = HS,
    Shots_Against = AS,
    ShotsOnTarget_For = HST,
    ShotsOnTarget_Against = AST,
    Corners_For = HC,
    Corners_Against = AC,
    Yellows_For = HY,
    Yellows_Against = AY,
    Reds_For = HR,
    Reds_Against = AR,
    Result = FTR
  ) %>%
  mutate(Home_Away = "Home")
# away
away_results <- match_data_scaled %>%
  select(Date, AwayTeam, FTAG, FTHG, AS, HS, AST, HST, AC, HC, AY, HY, AR, HR, FTR) %>%
  rename(
    Team = AwayTeam,
    Goals_For = FTAG,
    Goals_Against = FTHG,
    Shots_For = AS,
    Shots_Against = HS,
    ShotsOnTarget_For = AST,
    ShotsOnTarget_Against = HST,
    Corners_For = AC,
    Corners_Against = HC,
    Yellows_For = AY,
    Yellows_Against = HY,
    Reds_For = AR,
    Reds_Against = HR,
    Result = FTR
  ) %>%
  mutate(Home_Away = "Away")
# full history
team_results <- bind_rows(home_results, away_results) %>%
  arrange(Team, Date)
```

Create Weighted Past Features

```{r}
# more recent games are given a higher weight
weights <- rev(1:10)

# get weighted stats
get_weighted_stats <- function(past_matches) {
  n_matches <- nrow(past_matches)
  # if 0  
  if (n_matches == 0) {
    return(tibble(
      weighted_goals_for = NA,
      weighted_goals_against = NA,
      weighted_shots_for = NA,
      weighted_shots_against = NA,
      weighted_shots_on_target_for = NA,
      weighted_shots_on_target_against = NA,
      weighted_corners_for = NA,
      weighted_corners_against = NA,
      weighted_yellows_for = NA,
      weighted_yellows_against = NA,
      weighted_reds_for = NA,
      weighted_reds_against = NA,
      weighted_goal_diff = NA,
      weighted_win_rate = NA
    ))
  }
  
  use_weights <- tail(weights, n_matches)
  
  tibble(
    weighted_goals_for = weighted.mean(past_matches$Goals_For, use_weights),
    weighted_goals_against = weighted.mean(past_matches$Goals_Against, use_weights),
    weighted_shots_for = weighted.mean(past_matches$Shots_For, use_weights),
    weighted_shots_against = weighted.mean(past_matches$Shots_Against, use_weights),
    weighted_shots_on_target_for = weighted.mean(past_matches$ShotsOnTarget_For, use_weights),
    weighted_shots_on_target_against = weighted.mean(past_matches$ShotsOnTarget_Against, use_weights),
    weighted_corners_for = weighted.mean(past_matches$Corners_For, use_weights),
    weighted_corners_against = weighted.mean(past_matches$Corners_Against, use_weights),
    weighted_yellows_for = weighted.mean(past_matches$Yellows_For, use_weights),
    weighted_yellows_against = weighted.mean(past_matches$Yellows_Against, use_weights),
    weighted_reds_for = weighted.mean(past_matches$Reds_For, use_weights),
    weighted_reds_against = weighted.mean(past_matches$Reds_Against, use_weights),
    weighted_goal_diff = weighted.mean(past_matches$Goals_For - past_matches$Goals_Against, use_weights),
    weighted_win_rate = weighted.mean(ifelse(past_matches$Result == "H" & 
                                               past_matches$Home_Away == "Home", 1,
                                              ifelse(past_matches$Result == "A" & 
                                                       past_matches$Home_Away == "Away", 1, 
                                                     ifelse(past_matches$Result == "D", 0.5, 0))), use_weights)
  )
}
```

Loop Over Game and attach what we just added

```{r}
# initalize
home_features_list <- list()
away_features_list <- list()

for (i in 1:nrow(match_data_scaled)) {
  match <- match_data_scaled[i, ]
  # home
  home_past <- team_results %>%
    filter(Team == match$HomeTeam, Date < match$Date) %>%
    arrange(desc(Date)) %>%
    head(10)
  home_features <- get_weighted_stats(home_past)
  # away
  away_past <- team_results %>%
    filter(Team == match$AwayTeam, Date < match$Date) %>%
    arrange(desc(Date)) %>%
    head(10)
  away_features <- get_weighted_stats(away_past)
  
  # attach
  home_features_list[[i]] <- home_features
  away_features_list[[i]] <- away_features
}
# home dataset
home_features_df <- bind_rows(home_features_list) %>%
  rename_with(~ paste0("Home_", .x))
# away dataset
away_features_df <- bind_rows(away_features_list) %>%
  rename_with(~ paste0("Away_", .x))
# final dataset
model_data <- bind_cols(match_data_scaled, home_features_df, away_features_df)
```

*MODELING START*

Predicting Full Time Result (FTR):

Set Seed for Reproducability

```{r}
set.seed(486)
```

Dropping NA Values, believe this is because at start of season 10 games have not been completed yet

```{r}
model_data <- model_data %>%
  drop_na(
    B365H, B365D, B365A,
    norm_prob_H, norm_prob_D, norm_prob_A,
    Home_weighted_goals_for, Home_weighted_goals_against,
    Away_weighted_goals_for, Away_weighted_goals_against,
    Home_weighted_goal_diff, Away_weighted_goal_diff,
    Home_weighted_win_rate, Away_weighted_win_rate
  )
```

Chronoligcally Split Data

```{r}
train_data <- model_data %>% filter(Date < as.Date("2023-08-01"))
test_data  <- model_data %>% filter(Date >= as.Date("2023-08-01"))
```

Logistic Regression (Baseline), Has to be multinomial cuz 3 possible outcomes

```{r}
logit_ftr_model <- train(
  FTR ~ B365H + B365D + B365A +
  norm_prob_H + norm_prob_D + norm_prob_A +
  Home_weighted_goals_for + Home_weighted_goals_against +
  Away_weighted_goals_for + Away_weighted_goals_against +
  Home_weighted_goal_diff + Away_weighted_goal_diff +
  Home_weighted_win_rate + Away_weighted_win_rate +
  Home_weighted_shots_for + Home_weighted_shots_against +
  Away_weighted_shots_for + Away_weighted_shots_against +
  Home_weighted_shots_on_target_for + Away_weighted_shots_on_target_for +
  Home_weighted_corners_for + Away_weighted_corners_for,
  data = train_data,
  method = "multinom",
  trControl = trainControl(method = "none"),
  trace = FALSE
)
# preds
ftr_preds <- predict(logit_ftr_model, newdata = test_data)
# confusion matrix
confusionMatrix(ftr_preds, test_data$FTR)
```

LASSO Model

```{r}
lasso_ftr_model <- train(
  FTR ~ B365H + B365D + B365A +
  norm_prob_H + norm_prob_D + norm_prob_A +
  Home_weighted_goals_for + Home_weighted_goals_against +
  Away_weighted_goals_for + Away_weighted_goals_against +
  Home_weighted_goal_diff + Away_weighted_goal_diff +
  Home_weighted_win_rate + Away_weighted_win_rate +
  Home_weighted_shots_for + Home_weighted_shots_against +
  Away_weighted_shots_for + Away_weighted_shots_against +
  Home_weighted_shots_on_target_for + Away_weighted_shots_on_target_for +
  Home_weighted_corners_for + Away_weighted_corners_for,
  data = train_data,
  method = "glmnet",
  family = "multinomial",
  trControl = trainControl(method = "cv", number = 5),
  tuneLength = 10
)
# preds
ftr_preds_lasso <- predict(lasso_ftr_model, newdata = test_data)
# confusion matrix
confusionMatrix(ftr_preds_lasso, test_data$FTR)
```

XGBoost Model

```{r}
xgb_ftr_model <- train(
  FTR ~ B365H + B365D + B365A +
  norm_prob_H + norm_prob_D + norm_prob_A +
  Home_weighted_goals_for + Home_weighted_goals_against +
  Away_weighted_goals_for + Away_weighted_goals_against +
  Home_weighted_goal_diff + Away_weighted_goal_diff +
  Home_weighted_win_rate + Away_weighted_win_rate +
  Home_weighted_shots_for + Home_weighted_shots_against +
  Away_weighted_shots_for + Away_weighted_shots_against +
  Home_weighted_shots_on_target_for + Away_weighted_shots_on_target_for +
  Home_weighted_corners_for + Away_weighted_corners_for,
  data = train_data,
  method = "xgbTree",
  trControl = trainControl(method = "cv", number = 3),
  tuneLength = 3
)
```
```{r}
# seoerate shunks because of all the warning with xgBoost
# preds
ftr_preds_xgb <- predict(xgb_ftr_model, newdata = test_data)
# confusion matrix
confusionMatrix(ftr_preds_xgb, test_data$FTR)
```

Random Forest Model

```{r}
rf_ftr_model <- train(
  FTR ~ B365H + B365D + B365A +
  norm_prob_H + norm_prob_D + norm_prob_A +
  Home_weighted_goals_for + Home_weighted_goals_against +
  Away_weighted_goals_for + Away_weighted_goals_against +
  Home_weighted_goal_diff + Away_weighted_goal_diff +
  Home_weighted_win_rate + Away_weighted_win_rate +
  Home_weighted_shots_for + Home_weighted_shots_against +
  Away_weighted_shots_for + Away_weighted_shots_against +
  Home_weighted_shots_on_target_for + Away_weighted_shots_on_target_for +
  Home_weighted_corners_for + Away_weighted_corners_for,
  data = train_data,
  method = "rf",
  trControl = trainControl(method = "cv", number = 5),
  tuneLength = 5
)
```
```{r}
# preds
ftr_preds_rf <- predict(rf_ftr_model, newdata = test_data)
# confusion matrix
confusionMatrix(ftr_preds_rf, test_data$FTR)
```

Compare Results

```{r}
# Compute accuracy manually for each model
acc_rf_ftr    <- mean(ftr_preds_rf    == test_data$FTR)
acc_xgb_ftr   <- mean(ftr_preds_xgb   == test_data$FTR)
acc_lasso_ftr <- mean(ftr_preds_lasso == test_data$FTR)
acc_log_ftr   <- mean(ftr_preds   == test_data$FTR)

results <- data.frame(
  Model    = c("Random Forest", "XGBoost", "Lasso", "Logistic Regression"),
  Accuracy = c(acc_rf_ftr, acc_xgb_ftr, acc_lasso_ftr, acc_log_ftr)
)

# Plot using ggplot2
ggplot(results, aes(x = Model, y = Accuracy, fill = Model)) +
  geom_col(width = 0.6) +
  ylim(0, 1) +
  geom_text(aes(label = round(Accuracy, 3)), vjust = -0.5, size = 4.5) +
  labs(
    title = "Model Accuracy Comparison",
    y = "Accuracy",
    x = ""
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Predicting Over/Under 2.5 Goals:

Logistic Regression (Baseline)

```{r}
logit_ou_model <- train(
  OU_2.5 ~ B365H + B365D + B365A +
  norm_prob_H + norm_prob_D + norm_prob_A +
  Home_weighted_goals_for + Home_weighted_goals_against +
  Away_weighted_goals_for + Away_weighted_goals_against +
  Home_weighted_goal_diff + Away_weighted_goal_diff +
  Home_weighted_win_rate + Away_weighted_win_rate +
  Home_weighted_shots_for + Home_weighted_shots_against +
  Away_weighted_shots_for + Away_weighted_shots_against +
  Home_weighted_shots_on_target_for + Away_weighted_shots_on_target_for +
  Home_weighted_corners_for + Away_weighted_corners_for,
  data = train_data,
  method = "multinom",
  trControl = trainControl(method = "none"),
  trace = FALSE
)
# preds
ou_preds <- predict(logit_ou_model, newdata = test_data)

# confusion matrix
confusionMatrix(ou_preds, test_data$OU_2.5)
```

LASSO Model

```{r}
lasso_ou_model <- train(
  OU_2.5 ~ B365H + B365D + B365A +
  norm_prob_H + norm_prob_D + norm_prob_A +
  Home_weighted_goals_for + Home_weighted_goals_against +
  Away_weighted_goals_for + Away_weighted_goals_against +
  Home_weighted_goal_diff + Away_weighted_goal_diff +
  Home_weighted_win_rate + Away_weighted_win_rate +
  Home_weighted_shots_for + Home_weighted_shots_against +
  Away_weighted_shots_for + Away_weighted_shots_against +
  Home_weighted_shots_on_target_for + Away_weighted_shots_on_target_for +
  Home_weighted_corners_for + Away_weighted_corners_for,
  data = train_data,
  method = "glmnet",
  family = "multinomial",
  trControl = trainControl(method = "cv", number = 5),
  tuneLength = 10
)
# preds
ou_preds_lasso <- predict(lasso_ou_model, newdata = test_data)
# confusion matrix
confusionMatrix(ou_preds_lasso, test_data$OU_2.5)
```

XGBoost Model

```{r}
xgb_ou_model <- train(
  OU_2.5 ~ B365H + B365D + B365A +
  norm_prob_H + norm_prob_D + norm_prob_A +
  Home_weighted_goals_for + Home_weighted_goals_against +
  Away_weighted_goals_for + Away_weighted_goals_against +
  Home_weighted_goal_diff + Away_weighted_goal_diff +
  Home_weighted_win_rate + Away_weighted_win_rate +
  Home_weighted_shots_for + Home_weighted_shots_against +
  Away_weighted_shots_for + Away_weighted_shots_against +
  Home_weighted_shots_on_target_for + Away_weighted_shots_on_target_for +
  Home_weighted_corners_for + Away_weighted_corners_for,
  data = train_data,
  method = "xgbTree",
  trControl = trainControl(method = "cv", number = 3),
  tuneLength = 3
)
```
```{r}
# seperate because of all the warnings
# preds
ou_preds_xgb <- predict(xgb_ou_model, newdata = test_data)
# confusion matrix
confusionMatrix(ou_preds_xgb, test_data$OU_2.5)
```

Random Forest Model

```{r}
rf_ou_model <- train(
  OU_2.5 ~ B365H + B365D + B365A +
  norm_prob_H + norm_prob_D + norm_prob_A +
  Home_weighted_goals_for + Home_weighted_goals_against +
  Away_weighted_goals_for + Away_weighted_goals_against +
  Home_weighted_goal_diff + Away_weighted_goal_diff +
  Home_weighted_win_rate + Away_weighted_win_rate +
  Home_weighted_shots_for + Home_weighted_shots_against +
  Away_weighted_shots_for + Away_weighted_shots_against +
  Home_weighted_shots_on_target_for + Away_weighted_shots_on_target_for +
  Home_weighted_corners_for + Away_weighted_corners_for,
  data = train_data,
  method = "rf",
  trControl = trainControl(method = "cv", number = 5),
  tuneLength = 5
)
```
```{r}
# Predictions
ou_preds_rf <- predict(rf_ou_model, newdata = test_data)

# Confusion Matrix
confusionMatrix(ou_preds_rf, test_data$OU_2.5)
```

Compare Results

```{r}
# Compute accuracy manually for each model
acc_rf    <- mean(ou_preds_rf    == test_data$OU_2.5)
acc_xgb   <- mean(ou_preds_xgb   == test_data$OU_2.5)
acc_lasso <- mean(ou_preds_lasso == test_data$OU_2.5)
acc_log   <- mean(ou_preds   == test_data$OU_2.5)

results <- data.frame(
  Model    = c("Random Forest", "XGBoost", "Lasso", "Logistic Regression"),
  Accuracy = c(acc_rf, acc_xgb, acc_lasso, acc_log)
)

# Plot using ggplot2
ggplot(results, aes(x = Model, y = Accuracy, fill = Model)) +
  geom_col(width = 0.6) +
  ylim(0, 1) +
  geom_text(aes(label = round(Accuracy, 3)), vjust = -0.5, size = 4.5) +
  labs(
    title = "Model Accuracy Comparison",
    y = "Accuracy",
    x = ""
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```